Foraging: The Triumphant Return

### 2022-07-08 manuscript.tex
A few minor edits

Different aspects of foraging behaviour, such as the search image and the search path, clearly interact with one another, but most research has tried to understand them separately. In many cases, the pattern of foraging behaviour has also been studied using aggregate measures calculated on the level of a trial, such as the mean number of 'runs' (where the same type of target is selected multiple times in a row) or the total number of targets found during the longest run. However, we have recently developed a new model to analyse foraging data, based on a sampling without replacement procedure \cite{clarke2022foraging}. By using this model, we demonstrated that we were able to break down foraging into a number of different cognitive biases, such as a preference to stick to the same target type, or a preference for nearby targets, and used this model to successfully re-analyse data from a number of open access datasets \cite{kristjansson2014, thornton2020, clarke2022, tagu_kristjansson_2021}.

While our model is able to do a good job of allowing us to understand biases at the level of a trial (e.g. conjunction vs. feature search differences \cite{kristjansson2014} or the influence of high value targets compared to low value targets \cite{tagu2020}), we did not originally investigate to what extent our model was able to predict target-by-target behaviour in any given trial. However, as the model is based on target-by-target level information, it is possible for us to use this information to ask: to what extent does our model predict foraging behaviour within a trial? Can it predict exactly which target a participant will pick next? And if the model makes mistakes, can we understand where it is failing, in order to improve our understanding of how human foraging behaviour operates?

@@

We will a previously published dataset to explore how well our model can account for trial-level behaviour. We will start with the visual foraging data from \cite{clarke2022}. This is an attractive dataset for our needs as it is very close to the ``classic'' visual foraging paradigm \citep{kristjansson2014}, yet with a substantially larger sample of participants. (The key difference between paradigms is that \cite{kristjansson2014} used an ipad and finger foraging, while \cite{clarke2022} used a desktop computer and mouse clicks). Furthermore, we have already demonstrated that our model can capture individual differences in foraging behaviour \citep{clarke2022}. As such, we will only give a brief overview of the data here and refer the reader to these earlier papers for a more details.

@@

where $t_i = 1$ if item $i$ is of class $A$ and 0 otherwise, $m(t_i, t_{i-1}) =1$ if item $i$ is the same class as the previously selected item. $\rho_d$ and $\rho_{\theta}$ measure proximity and the effect of direction:

@@

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from \cite{clarke2022}. For each target selection, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next, easily outperforming a chance $1/n$ baseline. We can also see  that the model is well calibrated in that the probabilities assigned to the largest target manage to capture how well often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/qjep_preds.pdf}
\caption{(\textit{left}) Posterior probabilities for target selections during a visual foraging task. Each dot shows the  data from an individual participant, averaged over trials, in a condition and the shaded region indicates the interval in which we expect 67\% of participants to fall. The dashed line indicates chance performance. (\textit{centre})  Calibration plot for our foraging model. The $x$-axis gives the largest weight assigned by the model while the $y$-axis shows how often that target was actually selected by a human participant. (\textit{right}) This plot shows how often the 2nd and 3rd ranked items are selected based on the weights assigned by the model.}
\label{fig:cal}
\end{figure}

@@

One weakness of the foraging model by \cite{clarke2022foraging} is that it makes little attempt to predict which item will be selected first. In this section we aim to improve this by modelling participant bias (or preference) in their choice of initial target selection. Our approach is inspired by previous work on the central bias in scene viewing \cite{clarke_tatler2014, clarke2017}. These studies characterised fit truncated Gaussian distributions to the  $(x,y)$ coordinates of fixation locations. Here we will do similar, and fit beta distributions\footnote{We use beta distributions here as ADFC couldn't work out the syntax for specifying a mixture model using truncated Gaussians in Stan. But, in hindsight, beta distributions are probably a better choice anyway!} to locations of the initial target selections in the visual foraging task.

@@

From inspection of Figure \ref{fig:qjep_paths}, we noticed that in some cases, participants may be using a type of local path minimisation procedure, selecting a further away target in order to minimise overall path length, which the model does not always seem to predict. Could this help to explain some of the mistakes the model makes? Overall, the model doesn't seem to make substantially longer paths through the items than the human participants (see \textit{Supplementary Materials}). However, there is a relationship between model accuracy and the proportion of times that participants select paths shorter than the model: as model accuracy increases, people seem to be more likely to have paths shorter than the model predicts. This seems to suggest that path minimisation is not a good explanation for the targets our model is failing to capture, and instead the people that are difficult to predict are taking longer paths than the model thinks they should take.

Our exploratory analyses seem to hint at the idea that the people who are difficult for the model to predict are those who do not behave in a manner that optimises path length. It could be that these participants are simply unpredictable, and do not behave in a consistent manner on a target-by-target basis, making it difficult for a model to predict their behaviour at this level of granularity. However, it is possible that there are other factors that we have not taken into account that could further help explain these discrepancies. For example, it is possible that relatively unpredictable participants use multiple rules and switch between them in ways that we are not currently capturing in the model. Some participants may also lose track of where they are more frequently (perhaps due to attentional lapses) and thus may show foraging patterns that are predictable apart from unpredictable discontinuities, which again may not be easy to model. Finally, the displays had artificial discontinuities (i.e. screen edges) which may lead to different behaviour in different participants: some may carry out the task as if they are reading, reaching the right hand edge and then swinging back to the left, whereas others may 'bounce' off an edge they have just reached. Our model may well be able to predict the latter, but could struggle with the former behaviour.

@@

One of the key benefits of computational models is that they allow us to rigorously test how well we are able to predict behaviour, and can provide insights into what factors we are not capturing that may have important influences on how participants complete a task. We have demonstrated that our foraging model \cite{clarke2022foraging} is not only able to predict behaviour on aggregate, but can also make reasonable predictions at the target-by-target level. It is particularly good if a participant has a strong proximity bias, and on 'feature' trials where participants are normally switching between target types fairly frequently. It is also good at predicting where participants will start on a trial. The model finds it more difficult to predict 'discontinuous' jumps, which could be caused by a range of factors: a more difficult foraging task (e.g. conjunction searches), local path length minimisation, inattention, or aspects of the physical search space (e.g. edges of the screen). However, it would be possible to extend the model to incorporate these factors e.g. by incorporating a heuristic that would allow for local path minimisation, or introducing time as a predictor in the model. Our findings also suggest possible future directions for empirical work, such as evaluating the effect of inattention on foraging behaviour, or how the foraging targets are organised in space. We suggest that computational modelling is a powerful tool for helping us to understand behaviour, both by incorporating previous research into a shared framework, and by making testable predictions for future work.

>>>>

Different aspects of foraging behaviour, such as the search image and the search path, clearly interact with one another, but most research has tried to understand them separately. In many cases, the pattern of foraging behaviour has been studied using aggregate measures calculated on the level of a trial, such as the mean number of 'runs' (where the same type of target is selected multiple times in a row) or the total number of targets found during the longest run. These summary statistics can provide a broad view of switching behaviour, but they can also be biased by the spatial distribution of target types, and by a forager's preference for one target or another. To address the need for a more precise set of measurements, we recently developed a new model to analyse foraging data, based on a sampling without replacement procedure \cite{clarke2022foraging}. By using this model, we demonstrated that we were able to break down foraging into a number of different cognitive biases, such as a preference to stick to the same target type, or a preference for nearby targets, and used this model to successfully re-analyse data from a number of open access datasets \cite{kristjansson2014, thornton2020, clarke2022, tagu_kristjansson_2021}.

While our model is able to do a good job of allowing us to understand biases at the level of a trial (e.g. conjunction vs. feature search differences \cite{kristjansson2014} or the influence of high value targets compared to low value targets \cite{tagu2020}), we did not originally investigate to what extent our model was able to predict target-by-target behaviour within any given trial. However, as the model is based on target-by-target level information, it is possible for us to use this information to ask: to what extent does our model predict foraging behaviour within a trial? Can it predict exactly which target a participant will pick next? And if the model makes mistakes, can we understand where it is failing, in order to improve our understanding of how human foraging behaviour operates?

@@

We use a previously published dataset to explore how well our model can account for trial-level behaviour: the visual foraging data from \cite{clarke2022}. This is an attractive dataset for our needs as it is very close to the ``classic'' visual foraging paradigm \citep{kristjansson2014}, yet with a substantially larger sample of participants. (The key difference between paradigms is that \cite{kristjansson2014} used an ipad and finger foraging, while \cite{clarke2022} used a desktop computer and mouse clicks.) Furthermore, we have already demonstrated that our model can capture individual differences in foraging behaviour \citep{clarke2022}. As such, we will only give a brief overview of the data here and refer the reader to these earlier papers for more details.

@@

where $t_i = 1$ if item $i$ is of class $A$ and 0 otherwise, and  $m(t_i, t_{i-1}) =1$ if item $i$ is the same class as the previously selected item. $\rho_d$ and $\rho_{\theta}$ measure proximity and the effect of direction:

@@

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from \cite{clarke2022}. For each target selection, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next, easily outperforming a chance $1/n$ baseline. We can also see  that the model is well calibrated in that the probabilities assigned to the largest target manage to capture how often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/qjep_preds.pdf}
\caption{(\textit{left}) Posterior probabilities for target selections during a visual foraging task. Each dot shows the  data from an individual participant, averaged over trials, in a condition and the shaded region indicates the interval in which we expect 67\% of participants to fall. The dashed line indicates chance performance. (\textit{centre})  Calibration plot for our foraging model. The $x$-axis gives the largest weight assigned by the model while the $y$-axis shows how often that target was actually selected by a human participant. The dashed line here and in the right plot is the identity line. (\textit{right}) This plot shows how often the 2nd and 3rd ranked items are selected based on the weights assigned by the model.}
\label{fig:cal}
\end{figure}

@@

One weakness of the foraging model by \cite{clarke2022foraging} is that it makes little attempt to predict which item will be selected first. In this section we aim to improve this by modelling participant bias (or preference) in their choice of initial target selection. Our approach is inspired by previous work on the central bias in scene viewing \cite{clarke_tatler2014, clarke2017}. These studies fit truncated Gaussian distributions to the  $(x,y)$ coordinates of fixation locations. Here we will do similar, and fit beta distributions\footnote{We use beta distributions here as ADFC couldn't work out the syntax for specifying a mixture model using truncated Gaussians in Stan. But, in hindsight, beta distributions are probably a better choice anyway!} to locations of the initial target selections in the visual foraging task.

@@

From inspection of Figure \ref{fig:qjep_paths}, we noticed that in some cases, participants may be using a type of local path minimisation procedure, selecting a further away target in order to minimise overall path length, which the model does not always seem to predict. Could this help to explain some of the mistakes the model makes? Overall, the model doesn't seem to make substantially longer paths through the items than the human participants (see \textit{Supplementary Materials}). However, there is a relationship between model accuracy and the proportion of times that participants select paths shorter than the model: as model accuracy increases, people seem to be more likely to have paths shorter than the model predicts. This seems to suggest that path minimisation is not a good explanation for the participants our model is failing to capture, and instead the people that are difficult to predict are taking longer paths than the model thinks they should take.

Our exploratory analyses seem to hint at the idea that the people who are difficult for the model to predict are those who do not behave in a manner that optimises path length. It could be that these participants are simply unpredictable, and do not behave in a consistent manner on a target-by-target basis, making it difficult for a model to predict their behaviour at this level of granularity. However, it is possible that there are other factors that we have not taken into account that could further help explain these discrepancies. For example, it is possible that relatively unpredictable participants use multiple rules and switch between them in ways that we are not currently captured in the model. Some participants may also lose track of where they are more frequently (perhaps due to attentional lapses) and thus may show foraging patterns that are predictable apart from unpredictable discontinuities, which again may not be easy to model. Finally, the displays had artificial discontinuities (i.e. screen edges) which may lead to different behaviour in different participants: some may carry out the task as if they are reading, reaching the right hand edge and then swinging back to the left, whereas others may 'bounce' off an edge they have just reached. Our model may well be able to predict the latter, but could struggle with the former behaviour.

@@

One of the key benefits of computational models is that they allow us to rigorously test how well we are able to predict behaviour, and can provide insights into what factors we are not capturing that may have important influences on how participants complete a task. We have demonstrated that our foraging model \cite{clarke2022foraging} is not only able to predict behaviour in aggregate, but can also make reasonable predictions at the target-by-target level. It is particularly good if a participant has a strong proximity bias, and on 'feature' trials where participants are normally switching between target types fairly frequently. It is also good at predicting where participants will start on a trial. The model finds it more difficult to predict 'discontinuous' jumps, which could be caused by a range of factors: a more difficult foraging task (e.g. conjunction searches), local path length minimisation, inattention, or aspects of the physical search space (e.g. edges of the screen). However, it would be possible to extend the model to incorporate these factors e.g. by incorporating a heuristic that would allow for local path minimisation, or introducing time as a predictor in the model. Our findings also suggest possible future directions for empirical work, such as evaluating the effect of inattention on foraging behaviour, or how the foraging targets are organised in space. We suggest that computational modelling is a powerful tool for helping us to understand behaviour, both by incorporating previous research into a shared framework, and by making testable predictions for future work.
